Early evolution of the ecdysozoan body plan

Version 1

Abstract

Abstract

Extant ecdysozoans (moulting animals) are represented by a great variety of vermiform or articulated organisms. However, controversies remain about the nature of their ancestral body plan although the vermiform hypothesis seems to prevail. We describe here Beretella spinosa gen et sp. nov. a tiny ecdysozoan from the early Cambrian, Yanjiahe Formation, South China, with an unusual sack-like appearance, single opening, and spiny ornament. Beretella has no equivalent among animals, except Saccorhytus from the basal Cambrian. Phylogenetic analyses resolve both forms as a sister group (Saccorhytida) to all known Ecdysozoa, thus suggesting that ancestral ecdysozoans may have been non-vermiform animals. Saccorhytids are likely to represent an early dead-end off-shot along the stem-line Ecdysozoa that possibly evolved through anatomical simplification (e.g. lack of anus). Although extinct during the Cambrian, this animal lineage provides precious insight into the early evolution of Ecdysozoa and the nature (possibly non-vermiform) of the earliest representatives of the group.

Introduction

The Ediacaran–Cambrian transition is marked by the appearance in the fossil record of a variety of new body plans that prefigure the majority of present-day animal lineages, including the ecdysozoans a huge clade that encompasses all invertebrate animals growing through successive moulting stages, such as panarthropods (Arthropoda, Onychophora, Tardigrada), scalidophoran (incl.Priapulida) and nematoid worms(Erwin, 2020). Altogether ecdysozoans represent a very high percentage of animal biodiversity and disparity, inhabiting almost all possible ecological niches on Earth(Brusca et al., 2016). The nature of the last common ancestor of Ecdysozoa (LCAE) remains largely unresolved, even though worms are prevalent before the rise of panarthropods as trace and body fossils in basal Cambrian and late Ediacaran rocks(Buatois et al., 2014; Liu et al., 2014; Vannier et al., 2010). Some molecular phylogenies also predict that the most basal ecdysozoans were vermiform organisms(Howard et al., 2022; Laumer et al., 2019) that possibly diverged in the Ediacaran (Howard et al., 2022; Rota-Stabelli et al., 2013). Current reconstruction based on fossil and developmental evidence features the ancestral ecdysozoan as a millimeter-sized worm(Budd, 2001; Valentine and Collins, 2000) with a terminal(Ortega-Hernandez et al., 2019) or ventral mouth (Martín-Durán and Hejnol, 2015; Nielsen, 2019). Clearly, the discovery of Saccorhytus(Han et al., 2017; Liu et al., 2022; Shu and Han, 2020b) in the basal Cambrian of China (Kuanchuanpu Formation; ca. 535 Ma(Sawaki et al., 2008)) that is anything but a worm sowed doubt among scientists. Saccorhytus is a sac-like secondarily phosphatized microscopic animal spiked with conical sclerites and a single opening that was first seen as the earliest known deuterostome(Han et al., 2017) but is now considered as an ecdysozoan on more solid grounds(Liu et al., 2022; Shu and Han, 2020b), thus broadening the anatomical spectrum of the group and its disparity in the Cambrian and reopening the debate on the nature of LCAE.

We describe here Beretella spinosa gen. et sp. nov. from Member 5 of the Yanjiahe Formation (basal Cambrian Stage 2, ca. 529 Ma, Hubei Province, China) that shares morphological traits with Saccorhytus such as an ellipsoidal body, a pronounced bilaterality, a spiny ornament made of broad-based sclerites, and a single opening. Cladistic analyses are made to resolve the position of both Beretella and Saccorhytus that provide key information on the early evolution of the group.

Systematic palaeontology

Superphylum Ecdysozoa Aguinaldo et al.(Aguinaldo et al., 1997) Phylum Saccorhytida Han, Shu, Ou and Conway Morris, 2017 stat. nov.

Remarks

Saccorhytida first appeared in the literature as a new stem-group Deuterostomia that accommodated a single species, Saccorhytus coronarius(Han et al., 2017). Since Saccorhytus is no longer considered a primitive deuterostome and, instead, more likely belongs to ecdysozoans, Saccorhytida became an extinct Order of Ecdysozoa(Liu et al., 2022; Shu and Han, 2020b). Because both Saccorhytus and Beretella display major morphological differences with all other known ecdysozoan phyla (Nematoida, Scalidophora, and Panarthropoda), Saccorhytida is tentatively elevated here to the rank of phylum within Ecdysozoa.

Emended diagnosis

Microscopic, ellipsoidal body shape with pronounced bilateral symmetry expressed by paired spiny sclerites. Single, presumably oral opening on assumed ventral side (no anus).

Remarks

Only two forms, Saccorhytus and Beretella are currently placed within Saccorhytida, making it premature to formally define intermediate taxonomic categories such as an order and a family.

Beretella spinosa Han, Guo, Wang and Qiang, gen. et sp. nov.

Etymology

From ‘béret’, French, that designates a soft, visorless cap referring to the overall shape of this species, and ‘spinosa’, an adjective (Latin), alluding to its spiny ornament.

Holotype

CUBar138-12 (Fig. 1a–c).

Beretella spinosa gen. et sp. nov. from Member 5 of the Yanjiahe Formation (Cambrian Stage 2), Yichang, Hubei Province, China.

a–c, Holotype, CUBar138-12. a, Dorsal view showing the external ornament: (five sclerites at the midline in yellow (S1); flanked by two rows of sclerites in blue (S2); large broad-based conical sclerites in two dorsolateral pairs and one antero-posterior pairs in green (S3)); white arrows indicate lateral constriction. b, Right lateral view showing two additional rows of six sclerites (S4 and S5, in light blue and pink, respectively). c, Ventral view showing a large opening that may have accommodated the mouth (see the text) and an empty body cavity. d–g, CUBar75-45. d, Dorsal view showing a broken S3. e, Micro-CT image, right lateral view displaying S4. f, Ventral view depicting a tiny projection in purple. g, An enlargement of the projection of f. h–i, Paratype, CUBar171-5. h, Right dorsal view showing S1–S4. i, Right-lateral view showing S4 and S5. j–k, Paratype CUBar121-8. j, Dorsal view showing poorly preserved S1 and S2. k, Right-lateral view showing S3–S5. A, assumed anterior end (see text); ef, exotic fragment; D, assumed dorsal side; L, left; P, posterior end; R, right; tp, tiny spine; V, ventral side. The same abbreviations are used throughout the manuscript including Supplementary materials.

Paratype

CUBar171-5 Fig. 1h, i) and CURBar121-8 (Fig. 1j, k).

Diagnosis

Body with a beret-like lateral profile. Convex side (presumably dorsal) with an elevated (presumably posterior) and lower (presumably anterior one) end. The opposite side (presumably ventral) flattened. Bilateral symmetry well expressed in the overall body shape (sagittal plane) and sclerite distribution. Antero-posterior polarity. Convex side with a slightly elevated sagittal stripe topped with a single row of four aligned spines (S1) and five additional spines (S2) on each side. Six broad-based conical sclerites (S3) distributed in two symmetrical longitudinal rows plus two sagittal ones. Double rows of six marginal spines (S4 and S5). Flattened side often pushed in and partly missing, bearing a possible mouth opening. Possible oral spine.

Stratigraphy and locality

Watsonella crosbyi Assemblage Zone(Guo et al., 2021), Member 5 of the Yanjiahe Formation (Cambrian Terreneuvian, Stage 2) in the Yanjiahe section near Yichang City, Hubei Province, China (Supplementary Figs. 1, 2).

Description and comparisons

The body of Beretella spinosa is secondarily phosphatized and has a consistent beret-like three-dimensional shape in the lateral view. Its maximum length, width, and height range from 1.0–2.9 mm, to 975–2450 µm, and 500– 1000 µm, respectively (Fig. 1, Supplementary Tables 1–3). The ratio of the maximal length to width is 1.6:1 (Supplementary Fig. 3). As seen in top view, B. spinosa shows a small lateral constriction at approximately mid-length (Fig. 1a, c).

The body has a convex, assumedly dorsal side with one, presumably posterior end more elevated than the other (Fig. 1b, e, i, k). This elevation is gradual along the sagittal plane and then becomes more abrupt near the low elevated, presumably anterior end. The opposite, assumedly ventral side is less well preserved and seems to have been originally flattened.

The convex side bears a complex ornamented pattern made of five sets (S1– S5) of spiny sclerites directed towards the more elevated end (Figs. 1a, b, d, e, h–k, 2a, b, d). These sclerites were originally pointed (Figs. 1a, d, e, 2b, k, l, Supplementary Fig. 4a, b, g), but most of them were broken thus revealing an internal cavity and an ellipsoidal transverse section (Figs. 1a, b, h–k, 2a–e, g). The broken sclerites show an inner and outer phosphatic layer (thickness ca. 20 to 50 µm) often separated by a thin empty space (Fig. 2g–l).

Beretella spinosa gen. et sp. nov.

a, CUBar99-19, dorsal view showing an ornament S1–S4. b–c, CUBar136-9, general dorsal view and details. d, CUBar136-11, dorsal view showing S1–S5. e–f, CUBar73-15 general view and details of the cuticular polygonal reticulation in black. g–j, CUBar128-27. g–i, general view and details of the bi-layered structure of the cuticular wall as seen in broken conical sclerites. j, Micro-CT section showing possibly sclerite infilling. k-l, CUBar99-18, cuticular fragment, general view and details of large sclerite (central feature represents possible phosphatic infilling). is, infilling sclerite; pr, polygonal reticulation; sb, sclerite base.

The convex side bears six prominent conical sclerites (S3) all with a rounded to elliptical well-delimited broad base, distributed in two longitudinal symmetrical pairs with two additional sclerites at both ends of the sagittal plane (Figs. 1, 2d, Supplementary Figs. 3e–h, 4e–i). A low-relief stripe runs in a sagittal position and vanishes towards the elevated end. It is topped by a row of aligned spines (S1, Fig. 1a); the one closer to the more elevated end being more tubular and longer. This row is flanked on both sides by smaller aligned spines (S2, Figs. 1a, d, h, 2a–c). Two relatively sinuous rows of six tiny spines are present parallel to the lateral margins (S4 and S5, Figs. 1b, e, h–j, 2d, e). The convex side bears a polygonal micro-ornament (mesh size ca. 5 µm wide, Fig. 2f, Supplementary Tables 1–3). However, its exact extension is uncertain due to coarse secondary phosphatization. Clusters of spherical phosphatized grains (diameter ca. 20 µm) occur near the sclerite base (Supplementary Fig. 4b).

In most specimens, the flattened side is occupied by a relatively large opening (1200 and 600µm in maximal length and width, respectively) with irregularly defined margins (Fig. 1c, f, see also supplementary information movies 1, 2). The flattened side is often largely missing and opens into a spacious internal cavity with no signs of internal organs (e.g. gut and pharynx) (Fig. 1c, f). One specimen shows a tiny spine on the margin of the flattened side (Fig. 1f, g), which differs from other spiny sclerites (S1-S5).

The length of studied specimens ranges from 1.0 to 2.9 mm (Supplementary Fig. 3e–h). Whether growth was continuous or instead took place via successive moulting stages and cuticular renewal (ecdysis) could not be tested due to the small number of specimens (N=17) available for measurements. No major morphological variations (e.g. a sclerite pattern) can be seen between the smallest and largest specimens of B. spinosa (Supplementary Fig. 3e–h).

Remarks

Body polarities in Beretella

The anterior-posterior (AP) and dorsal-ventral (DV) polarities of Beretella are uneasy to define because of the lack of modern equivalent among extant animals. In the vast majority of extinct and extant invertebrates for which antero-posterior polarity is defined on the basis of independent criteria (e.g. position of mouth), sclerites point backwards (e.g. Cambrian scalidophoran worms (Han et al., 2007; Huang et al., 2004) and Wiwaxia (Zhang et al., 2015b)). This is most probably also the case with Beretella (Fig. 1a, d, j). The dorsoventral polarity of Beretella is supported by the fact that protective sclerites such as spines most commonly occur on the dorsal side of bilaterians (Fig. 1a, d, j).

Comparison with Saccorhytus and other ecdysozoans

Beretella spinosa has no exact equivalent in any Cambrian animals except Saccorhytus coronarius, an enigmatic, sac-like ecdysozoan(Han et al., 2017; Liu et al., 2022; Shu and Han, 2020b). Both forms share a tiny, poorly differentiated ellipsoidal body, and a set of prominent bilaterally arranged spiny sclerites. Indeed, the broad-based conical sclerites (S3) of Beretella are almost identical to those of Saccorhytus (Supplementary Fig. 4c) and have counterparts among scalidophoran worms (Supplementary Fig. 4d). However, they differ in number, ornamented structures, shape, and spatial arrangement (see details in Supplementary Tables 1–3) which makes the hypothesis of Saccorhytus being the larval stage of Beretella unlikely. Both Beretella and Saccorhytus differ from other known ecdysozoans in the lack of a vermiform body, introvert, annulations, and through gut (Supplementary Tables 1–3).

Discussion

Ventral mouth

All bilaterian animals have a digestive system with at least one opening that corresponds to the mouth(Brusca et al., 2016). Although the presumed oral area of Beretella is poorly preserved (ventral side often pushed in and largely destroyed), its mouth is likely to be found ventrally (see description), since no other opening occurs on its dorsal side, except those created by broken sclerites. The well-defined dorsoventral polarity of Beretella would suggest that the animal was resting on its ventral (flattened) side, the spiny dorsal side playing a protective role. Maintaining ventral contact with a substrate seems to be very unlikely unless these microscopic ellipsoidal animals were interstitial.

Phylogenetic position of Beretella

Beretella’s phylogenetic affinities remain elusive due to the lack of information concerning its internal anatomy and ventral side. Its scleritome consists of isolated conical sclerites that were the cuticular outgrowths of a seemingly rigid integument that covered both sides of the animal. Such conical sclerites have close counterparts in Cambrian ecdysozoans such as scalidophoran worms (e. g. Eokinorhynchus(Zhang et al., 2015a)), lobopodians (e.g. Onychodictyon ferox(Hou et al., 1991)) and even more clearly Saccorhytus that recent cladistic analyses resolved as a branch of the total-group Ecdysozoa (Liu et al., 2022). These sclerites unknown in other animal groups, suggest that both Saccorhytus and Beretella belongs to Ecdysozoa in the absence of more direct fossil evidence such as exuviae or features suggesting cuticular moulting(Daley and Drage, 2016; Wang et al., 2019).

Cladistic analyses were performed to test the relation of Beretella and Saccorhytus to other ecdysozoan groups and, more generally, their phylogenetic relationships with other bilaterian groups (see details in Supplementary Table 4). Both taxa join in a clade (Saccorhytida, Fig. 3a–c) that is resolved as stem species within total-group Ecdysozoa and as the sister group of Cycloneuralia plus Panarthropoda, i.e. crown-group Ecdysozoa (Figs. 3d, 4, Supplementary Figs. 6-9). These results are consistent with the body plan of Saccorhytida being markedly different from that of crown-group ecdysozoans that all have a vermiform body and differentiated structures such as the introvert and pharyngeal complex (Fig. 4).

Position of Beretella spinosa in the animal tree based on cladistic analysis.

a-c, artistic three-dimensional reconstructions of Beretella spinosa in the anterolateral, dorsal, and posterolateral views. d, Animal tree obtained from cladistic analyses using maximum likelihood tree obtained from cladistic analyses using maximal likelihood (IQTREE). Saccorhytus and Beretella join in a clade (new phylum Saccorhytida) resolved as the sister-group of all other ecdysozoans; numbers at key nodes denote probability. Fossil and extant taxa are in italics and bold, respectively. Known fossil record indicated by thicker bars (after (Shu and Han, 2020a)).

Possible evolutionary scenario to explain the origin and early evolution of ecdysozoans.

a, Summary tree (see Supplementary Figs. 6-9) showing saccorhytids as a sister-group of Cycloneuralia (Nematoida plus Scalidophora) + Panarthropoda; main morphological features of each group listed along each branch. b, Potential evolutionary pathway to evolve Saccorhytida and crown-group Ecdysozoa. Numbers in green, red and blue circles designate pre-ecdysozoan (Spiralia), Saccorhytida and Cycloneuralia, respectively. Light brown gradient (circle) to emphasize ecdysis and sclerite secretion seen as key evolutionary steps. 1, Hypothetical pre-ecdysozoan animal with a ciliated epidermis and glycocalyx. 2, Saccorhytid exemplified by Beretella with a cuticle bearing sclerites and a simplified internal organization (e.g. loss of anus). 3, Crown-group ecdysozoan exemplified by a scalidophoran worm with an elongated shape, a differentiated head (introvert) and trunk, sclerites, a through gut, a terminal mouth and abilities to burrow into bottom sediment. Animals not to scale. Abbreviations: a, anus; ci, cilia; cu, cuticle; ec, epidermal cell; gl, glycocalyx (mucous layer); m, mouth; in, introvert; sc, sclerite; se, sediment; TGE, total-group Ecdysozoa. Silhouettes from phylopic.org.

The ancestral ecdysozoan body plan

Molecular clock analyses often place the divergence of Ecdysozoa relatively deep into the Ediacaran (Howard et al., 2022; Rota-Stabelli et al., 2013), thus highlighting major discrepancy with the known fossil record of the group. Potential ecdysozoans occur in the late Precambrian as suggested by 1) sclerites resembling those of extant priapulids, found in Ediacaran Small Carbonaceous Fossils assemblages(Moczydłowska et al., 2015) and 2) locomotion traces presumably made by scalidophoran worms (Buatois et al., 2014; Vannier et al., 2010). In the absence of fossil data for other vermiform groups such as nematoids, scalidophorans are potentially the oldest known representatives of Ecdysozoa. Recent Bayesian analyses based on a large molecular data set obtained from the 8 extant ecdysozoan phyla recover Scalidophora as the sister-group to Nematoida + Panarthropoda and suggest that ecdysozoans probably diverged in the Ediacaran possibly some 23 million years before the oldest fossil occurrence (trace fossils) of the group (Howard et al., 2022). Although this study does not speculate on the nature of the last common ancestor of Ecdysozoa, it is consistent with the view that the earliest representatives of the group were probably vermiform. Howard et al. (Howard et al., 2022) drew comparable conclusions based on Acosmia, an assumed stem-ecdysozoan worm from early Cambrian Chengjiang Lagerstätte. However, the re-evaluation of the morphological characteristics of this worm rather suggests a less basal position either within the total-group Cycloneuralia (Fig. 3d, Supplementary Figs. 6, 7) or among crown-group Ecdysozoa (Supplementary Figs. 8, 9). The non-vermiform nature of saccorhytids and their position as the sister group of the crown-group Ecdysozoa clearly reopens the debate on the nature of the ancestral ecdysozoan (Fig. 4) and suggests exploring various evolutionary hypotheses, in particular: 1) does the enigmatic saccorhytid body plan results from anatomical simplification? 2) to what extent may these animals shed light on the nature of the earliest ecdysozoans?

Do saccorhytids result from simplification?

A relatively simple body plan and tiny size is often seen as resulting from anatomical simplification (e.g. digestive system) and miniaturization (micrometric size) in possible relation with the adaptation to specialized ecological niches or parasitism (Hanken and Wake, 1993). For example, some extant scalidophoran worms living in interstitial (meiobenthic) habitats such as loriciferans have a miniaturized body (Kirstensen, 1983) compared with their macroscopic counterparts (e.g. Priapulus(Schmidt-Rhaesa, 2013b)). However, they retain a through gut and a functional introvert and show no sign of drastic internal simplification (Schmidt-Rhaesa, 2013a). Anatomical reduction is a typical feature of parasitism(Hanken and Wake, 1993) that is well-represented among extant ecdysozoans such as nematodes(Schmidt-Rhaesa, 2014). Although relatively small (ca. 0.1-2.5 mm long), nematodes underwent no simplification of their digestive system. Saccorhytids have no specialized features (e.g. anchoring or piercing structures) that would point to any adaptation to ecto- or endo-parasitic lifestyles(Cong et al., 2017). Saccorhytus has been interpreted16 as a possible interstitial animal based on its micrometric size which corresponds to that the of the extant meiofauna. If we accept the hypothesis of saccorhytids resulting from simplification, then we need to determine its origin. Simplification of saccorhytids from a vermiform animal (e.g. cycloneuralian worm with a through gut and terminal mouth) is difficult to conceive because it would involve considerable anatomical transformations such as the loss of vermiform organization, introvert and pharynx in addition to that of the digestive system (Fig. 4). Alternative options to consider are ancestral and not necessarily vermiform ecdysozoans.

Early evolution of ecdysozoans: a new scenario

We propose here an alternative evolutionary hypothesis (Fig. 4) in which saccorhytids are replaced within the broader framework of the origin and early diversification of moulting animals. Saccorhytids are seen as an early off-shot from the stem-line Ecdysozoa (see cladistic analysis above) that possibly retained important features of the body plan of ancestral non-vermiform ecdysozoans (see ancestral character state reconstruction in Supplementary Table 4). This scenario must be considered as a working hypothesis whose aim is to stimulate research in this key area of animal evolution.

The cuticular secretion and the loss of cilia (Valentine and Collins, 2000) are seen as the first of a series of evolutionary events (Fig. 4) that led to the rise of Ecdysozoa. Moulting (shedding of the old cuticle via apolysis and its renewal) further reconciled body growth and cuticular protection (Schmidt-Rhaesa, 2007). Cuticle secretion and moulting may have been quasi-simultaneous innovations that took place over a relatively short time interval. The nature of the very first ecdysozoans is hypothetical and lacks fossil evidence. However, they are tentatively represented here as small epibenthic or interstitial slow-moving non-vermiform animals from which saccorhytids may have evolved via an assumed anatomical simplification (i.e. loss of anus seen details in Supplementary Table 4, Fig. 4).

In our scenario, this ancestral ecdysozoan stock would have also given rise to vermiform ecdysozoans through stepwise anatomical transformations such as the body elongation, the differentiation of key morpho-functional structures such as the pharynx and the introvert and the shift of the ventral mouth to a terminal position (Martín-Durán and Hejnol, 2015) (Fig. 4). This mouth shift from ventral to terminal arising in crown ecdysozoans is consistent with the chronology of divergence of animal lineages and the fact that the mouth of most spiralians is ventral (Martín-Durán and Hejnol, 2015; Nielsen, 2019; Ortega-Hernandez et al., 2019). Developmental studies show that embryos of extant cycloneuralians have a ventral mouth that moves to a terminal position towards the adult stage (Martín-Durán and Hejnol, 2015; Nielsen, 2019). These assumed major anatomical changes (e.g. functional introvert) must be placed in the ecological context of Cambrian animal radiation. Important changes in the functioning of marine ecosystems occurred in the early Cambrian such as interactive relationships between animal species, exemplified by predation (Vannier and Chen, 2005; Vermeij, 1977) may have acted as drivers in the evolution of early ecdysozoans, in promoting burrowing into sediment and the colonization of endobenthic habitats for the first time (Vannier et al., 2010). Burrowing into the sediment could be seen as the evolutionary response of epibenthic animals such as ancestral ecdysozoans to escape visual predation(Daley et al., 2013; Vannier and Chen, 2005). This migration to endobenthic shelters was made possible by the development of a resistant cuticular layer (Fig. 4) that strongly reduced physical damage caused by friction with the sediment and provided anchoring points (e.g. scalids and sclerites). Whereas saccorhytids became rapidly extinct during the Cambrian, worms massively colonized endobenthic habitats, resulting in bioturbation and ecological turnover.

Methods

Material

Fourteen specimens of Beretella spinosa were recovered from samples (siliceous-phosphatic, intraclastic limestone) collected from Member 5 of the Yanjiahe Formation, Yanjiahe section near Yichang City, Hubei Province, China(Guo et al., 2021) (Supplementary Tables 1–3). These were obtained by digesting the rocks in 10% acetic acid. Faunal elements associated with Beretella spinosa (Supplementary Tables 1–3) in residues are mainly tiny molluscs (CUBar21-4 and CUBar206-6). Comparisons were made with 10 specimens of Saccorhytus coronarius (ELIXX25-62, ELIXX34-298, ELIXX45-20, ELIXX48-64, ELIXX58-336, ELIXX61-27, ELIXX65-116, ELIXX65-296, ELIXX99-420) and one coeval scalidophoran specimen (ELIXX57-320) all from Bed 2 of the Kuanchuanpu Formation, Zhangjiagou section near Xixiang County, south Shaanxi Province, China. All specimens of Beretella are deposited in the paleontological collections of Chang’an University, Xi’an (CU), those of scalidophoran, and Saccorhytus at Northwest University, Xi’an (ELI), China.

Scanning electron microscopy (SEM)

All specimens were coated with gold and then imaged using a FEI Quanta 400 FEG SEM at Northwest University and a FEI Quanta 650 at Chang’an University.

X-ray computed microtomography and 3D reconstruction

Micro-CT-images (tiff format, with pixel size 1.1 µm) of Beretella (CUBar75-45, CUBar128-27, CUBar138-12) and Saccorhytus (ELIXX65-116, ELIXX99-420) were acquired using the Zeiss Xradia 520 at Northwest University (NWU), Xi’an, China, at an accelerating voltage of 50 kV and a beam current of 80 µA. Micro-CT data were processed using VGstudio Max 3.2 for 3D volume rendering.

Measurements

Measurements of the length, width, and height of Beretella and Saccorhytus were obtained from Micro-CT and SEM images by using tipDig2 v.2.16.

Phylogenetic analysis

We built our matrix with 55 taxa coded using 191 morphological characteristics (Supplementary Texts 1, 2). It is largely based on the data published by Howard et al.(Howard et al., 2020), Vinther and Parry(Vinther and Parry, 2019) and Ou et al.(Ou et al., 2017), although emended and supplemented by recent updates and new observations (Supplementary Text 1). Three characters (37. Through gut, 38. U-shaped gut, and 40. Ventral mouth) in matrix were coded as “? (uncertain)”, “?”, and “?”, respectively. Because although we can infer a ventral mouth and no anus of Beretella, these anatomic structures are invisible in fossils. We analyzed the data matrix using parsimony (TNT), likelihood (IQTREE) and Bayesian inference (MrBayes). Parsimony analysis was implemented in TNT under equal and implied (k=3) weight. Parameters are default (Goloboff et al., 2008; Goloboff and Catalano, 2016). The maximum-likelihood tree search was conducted in IQ-TREE(Nguyen et al., 2015), and support was assessed using the ultrafast phylogenetic bootstrap replication method (Hoang et al., 2018; Minh et al., 2013) to run 50,000 replicates. Bayesian inference was conducted in with MrBayes v3.2.6a with default priors and Markov chain Monte Carlo settings(Ronquist et al., 2012). Two independent runs of 7,000,000 Markov chain Monte Carlo generations were performed, each containing four Markov chains under the Mkv + Γ model for the discrete morphological character data(Lewis, 2001). In each run (N=2), trees were collected at a sampling frequency of every 5,000 generations and with the first 25% samples discarded as burn-in. The convergence of chains was checked by effective sample size (ESS) values over 1,000 in Tracer v.1.7(Rambaut et al., 2018), 1.0 for the potential scale reduction factor (PSRF)(Gelman and Rubin, 1992), and by an average standard deviation of split frequencies below 0.007.

Ancestral character state reconstructions

Ancestral character state reconstructions for six morphological characters were performed on the ecdysozoan total group node, the ecdysozoan crown group node and saccorhytid node. Characters selected for ancestral state reconstruction represent traits inferred as ecdysozoan plesiomorphies (ancestral characters) from studies of crown group taxa. These characters included the presence or absence of: (1) through gut; (2) ventral mouth; (3) introvert (see Supplementary Table 4).

This was carried out individually for the selected character in MrBayes. This was employed to calculate the posterior probability of the presence (1) and absence (0) of the selected characters at the selected nodes. Analyses used the MK + gamma model, and always converged after 2 million generations. Average deviation of split frequencies (< 0.01), ESS scores (> 200), and PSRF values (= approx. 1.00) assessed convergence of the MCMC chains (Howard et al., 2020).

eLife Assessment

This study provides a fundamental advance in palaeontology by reporting the fossils of a new invertrebrate, Beretella spinosa, and inferring its relationship with already described species. The analysis placed the newly described species in the earliest branch of moulting invertebrates. The study, supported by convincing fossil observation, hypothesizes that early moulting invertebrate animals were not vermiform.

Reviewer #1 (Public Review):

Summary:

Wang and co-workers characterise the fossil of Beretella spinosa from the early Cambrian, Yanjiahe Formation, South China. Combining morphological analyses with phylogenetic reconstructions, the authors conclude that B. spinosa is closely related to Saccorhytus, an enigmatic fossil recently ascribed to Ecdysozoa, or moulting animals, as an extinct "basal" lineage. Finding additional representatives of the clade Saccorhytida strengthens the idea that there existed a diversity of body plans previously underappreciated in Ecdysozoa, which may have implications for our understanding of the earliest steps in the evolution of this major animal group.

Strengths:

I'm not a paleobiologist; therefore, I cannot give an expert opinion on the descriptions of the fossils. However, the similarities with Saccorhytus seem evident, and the phylogenetic reconstructions are adequate. Evolutionary interpretations are generally justified, and the consolidation of Saccorhytida as the extinct sister lineage to extant Ecdysozoans will have significant implications for our understanding of this major animal clade.

Weaknesses:

While I generally agree with the author's interpretations, the idea of Saccorhytida as a divergent, simplified off-shot is slightly contradictory with a probably non-vermiform ecdysozoan ancestor. The author's analyses do not discard the possibility of a vermiform ecdysozoan ancestor (importantly, Supplementary Table 4 does not reconstruct that character), and outgroup comparison with Spiralia (and even Deuterostomia for Protostomia as a whole) indicates that a more or less anteroposteriorly elongated (i.e., vermiform) body is likely common and ancestral to all major bilaterian groups, including Ecdysozoa. Indeed, Figure 4b depicts the potential ancestor as a "worm". The authors argue that the simplification of Saccorhytida from a vermiform ancestor is unlikely "because it would involve considerable anatomical transformations such as the loss of vermiform organization, introvert, and pharynx in addition to that of the digestive system". However, their data support the introvert as a specialisation of Scalidophora (Figure 4a and Supplementary Table 4), and a pharyngeal structure cannot be ruled out in Saccorhytida. Likewise, loss of an anus is not uncommon in Bilateria. Moreover, this can easily become a semantics discussion (to what extent can an animal be defined as "vermiform"? Where is the limit?). Therefore, I suggest to leave the evolutionary scenario more open. Supporting Saccorhytida as a true group at the early steps of Ecdysozoa evolution is important and demonstrates that animal body plans are more plastic than previously appreciated. However, with the current data, it is unlikely that Saccorhytida represents the ancestral state for Ecdysozoa (as the authors admit), and a vermiform nature is not ruled out (and even likely) in this animal group. Suggesting that the ancestral Ecdysozoan might have been small and meiobenthic is perhaps more interesting and supported by the current data (phylogeny and outgroup comparison with Spiralia).

Reviewer #2 (Public Review):

Summary:

This work provides important anatomical features of a new species from the Lower Cambrian, which helps advance our understanding of the evolutionary origins of animal body plans. The authors interpreted that the new species possessed a bilateral body covered with cuticular polygonal reticulation and a ventral mouth. Based on cladistic analyses using maximum likelihood, Bayesian, and parsimony, the new species was placed, along with Saccorhytus, in a sister group ("Saccorhytida") of the Ecdysozoa. The phylogenetic position of Saccorhytida suggests a new scenario of the evolutionary origin of the crown ecdysozoan body plan.

Strengths:

Although the new species reported in this paper show strange morphologies, the interpretation of anatomical features was based on detailed observations of multiple fossil specimens, thereby convincing at the moment. Morphological data about fossil taxa in the Ediacaran and Early Cambrian are quite important for our understanding of the evolution of body plans (and origins of phyla) in paleontology and evolutionary developmental biology, and this paper represents a valuable contribution to such research fields.

Weaknesses:

The preservations of the specimens, in particular on the putative ventral side, are not good, and the interpretation of the anatomical features needs to be tested with additional specimens in the future. The monophyly of Cycloneuralia (Nematoida + Scalidophora) was not necessarily well-supported by cladistic analyses, and the evolutionary scenario (Figure 4) also needs to be tested in future works.

Reviewer #3 (Public Review):

Summary:

The authors of this manuscript identified the fossils of the newly designated species Beretella spinosa and analyzed its phylogenetic position in relation to the extinct described species and extant species. Their analysis placed the newly described species Beretella spinosa and Saccorhytus as an independent clade from the rest of the ecdysozoans. Remarkably, these species are non-vermiform, and the resulting evolutionary scenario assumes non-vermiform as early ecdysozoans.

Strengths:

The study presents outstanding, novel data and provides new insights into the evolution of animal forms especially regarding their morphological diversity after the Cambrian explosion.

Weaknesses:

I, as a paleontology non-expert, experienced several difficulties in reading the manuscript. This should be taken into consideration when assuming a wide range of readers including non-experts.

Author response:

The following is the authors’ response to the original reviews.

Public reviews:

Reviewer 1:

Weaknesses:

While I generally agree with the author's interpretations, the idea of Saccorhytida as a divergent, simplified off-shot is slightly contradictory with a probably non-vermiform ecdysozoan ancestor. The author's analyses do not discard the possibility of a vermiform ecdysozoan ancestor (importantly, Supplementary Table 4 does not reconstruct that character),

Saccorhytids are only known from the early Cambrian and their unique morphology has no equivalent among any extinct or extant ecdysozoan groups. This prompted us to consider them as a possible dead-end evolutionary off-shot. The nature of the last common ancestor of ecdysozoan (i.e. an elongated worm-like or non-vermiform animal with capacities to renew its cuticle by molting) remains hypothetical. At present, palaeontological data do not allow us to resolve this question. The animal in Fig. 4b at the base of the tree is supposed to represent an ancestral soft-bodied form with no cuticle from which ecdysozoan evolved via major innovations (cuticular secretion and ecdysis). Its shape is hypothetical as indicated by a question mark. Our evolutionary model is clearly intended to be tested by further studies and hopefully new fossil discoveries.

…and outgroup comparison with Spiralia (and even Deuterostomia for Protostomia as a whole) indicates that a more or less anteroposteriorly elongated (i.e., vermiform) body is likely common and ancestral to all major bilaterian groups, including Ecdysozoa. Indeed, Figure 4b depicts the potential ancestor as a "worm". The authors argue that the simplification of Saccorhytida from a vermiform ancestor is unlikely "because it would involve considerable anatomical transformations such as the loss of vermiform organization, introvert, and pharynx in addition to that of the digestive system". However, their data support the introvert as a specialisation of Scalidophora (Figure 4a and Supplementary Table 4), and a pharyngeal structure cannot be ruled out in Saccorhytida. Likewise, loss of an anus is not uncommon in Bilateria. Moreover, this can easily become a semantics discussion (to what extent can an animal be defined as "vermiform"? Where is the limit?).

We agree that “worm” and “vermiform” are ill-defined terms. They are widely used in various palaeontological and biological papers to describe elongated tubular animals such as edydsozoans and annelids (see Giribet and Edgecombe 2017; popular textbook written by Nielsen 2012; Schmit-Rhaesa 2013; Brusca et al. 2023; Giribet and Edgecombe 2020). Very few other animals are termed “worms”. Changes have been made in the text to solve this semantic problem, for example in the abstract where we added (i.e elongated and tubular) to better define what we mean by “vermiform”.

Priapulid worms or annelids are examples of extremely elongated, tubular animals. In saccorhytids, the antero-posterior elongation is present (as it is in the vast majority of bilaterians) but extremely reduced, Saccorhytus and Beretella having a sac-like or beret-shape, respectively. That such forms may have derived from elongated, tubular ancestors (e.g. comparable with present-day priapulid worms) would require major anatomical transformations that have no equivalent among modern animals. We agree that further speculation about the nature of these transformations is unnecessary and should be deleted simply because the nature of these ancestors is purely hypothetical. We also agree that the loss of anus and the extreme simplification of the digestive system is common among extant bilaterians. In Figure 4b, the hypothetical pre-ecdysozoan animal is slightly elongated (along its antero-posterior axis) but in no way comparable with a very elongated and cylindrical ecdysozoan worm (e.g. extant or extinct priapulid).

Therefore, I suggest to leave the evolutionary scenario more open. Supporting Saccorhytida as a true group at the early steps of Ecdysozoa evolution is important and demonstrates that animal body plans are more plastic than previously appreciated. However, with the current data, it is unlikely that Saccorhytida represents the ancestral state for Ecdysozoa (as the authors admit), and a vermiform nature is not ruled out (and even likely) in this animal group. Suggesting that the ancestral Ecdysozoan might have been small and meiobenthic is perhaps more interesting and supported by the current data (phylogeny and outgroup comparison with Spiralia).

We agree to leave the evolutionary scenario more open, especially the evolutionary process that gave rise to Saccorhytida. Again, we know nothing about the morphology of the ancestral ecdysozoan (typically the degree of body elongation, whether it had a differentiated introvert or not, whether it had a through gut or not). In Fig.4, the ancestral ecdysozoan is supposed to have evolved from a soft-bodied epibenthic animal through key innovations such as the secretion of a cuticle and ecdysis. It is a hypothesis that needs to be tested by further studies and fossil discoveries. Speculations concerning the process through which saccorhytids may have arisen have been deleted.

Reviewer 2:

Weaknesses:

The preservations of the specimens, in particular on the putative ventral side, are not good, and the interpretation of the anatomical features needs to be tested with additional specimens in the future. The monophyly of Cycloneuralia (Nematoida + Scalidophora) was not necessarily well-supported by cladistic analyses, and the evolutionary scenario (Figure 4) also needs to be tested in future works.

Yes, we agree that the animal described in our manuscrip remains enigmatic (e.g. the natures of its internal organs, its lifestyle, etc..). Whereas the dorsal side of the animal is well documented (consistent pattern of pointed sclerites), uncertainties remain concerning its ventral anatomy (typically the mouth location and shape). Additional better-preserved specimens will hopefully provide the missing information. Concerning Cycloneuralia, their monophyly is generally better supported by analyses based on morphological characters than in molecular phylogenies.

Reviewer 3:

Weaknesses:

I, as a paleontology non-expert, experienced several difficulties in reading the manuscript. This should be taken into consideration when assuming a wide range of readers including non-experts.

We have ensured that the text is comprehensible to biologists. The main results are summarized in relatively simple diagrams (e.g. Fig. 4) that can be understood by non-specialized readers. We are aware that technical descriptive terms may appear obscure to non-specialists. We can hardly avoid them in the descriptive parts. However, our figures (e.g. SEM images and 3D-reconstruction) are clear enough to give the reader a clear idea of the morphology of Beretella.

Recommendations for the authors:

All three reviewers appreciate the discovery and found the merit of publishing this manuscript. They also raised some concerns about the data presentation. The authors are requested to perform no additional analysis but to go through all the reviewer comments and rebut or intake them in revising the manuscript.

Reviewer 1:

- Line 41: comma after "ecdysozans".

OK, done.

- Formatting style: add a space before references.

OK, done.

- Line 169: B. spinosa in italics

OK, done.

- Line 157: could the "relatively large opening" in the flattened ventral side of a mouth (even when altered by the fossilisation process)?

Most bilaterians have a mouth. There is no opening on the relatively well-preserved dorsal side of Beretella, that could be interpreted as a mouth. In contrast the flattened ventral side often show a depressed area that could potentially bear a mouth. This ventral area is often pushed in and poorly preserved. The cuticle of this ventral side might have been relatively thinner, perhaps more flexible than that of the dorsal one (with strong sclerites). These differences might explain why the possible oral area is poorly preserved.

- Line 178: "position of the mouth"

OK, done.

- Line 219: "These sclerites, unknown..."

OK, done.

- Line 282: update reference formatting

OK, done.

- Line 298: remove reference to Supplementary Table 4, as it does not refer to the possible vermiform nature of the last common ecdysozoan ancestor?

OK, done.

- Figure 4a: change "paired legs" for "paired appendages"?

OK, done.

- Supplementary Table 4: For TGE and Introvert, the state 0 (absent) should be in bold and underlined (as it is the most likely state).

OK, done.

Reviewer 2:

Line 25: "from the early Cambrian" should be changed into "from the lower Cambrian"

OK, done.

Line 126: The range of maximum length should be reported in µm (rather than mm) just like those of maximum width and height.

OK, done.

Lines 191-192: Please recheck the figure panels of Saccorhytus (Supplementary Figure 4c) and scalidophoran worm (Supplementary Figure 4d). Perhaps, the former should refer to Figure 4d, and the latter to Figure 4c?

OK, done.

Lines 239 and 241: "1" and "2" appear to stand for citations (the other journal style), but I am not certain what they are.

To avoid confusing, we replace ‘1’ and ‘2’ by ‘i’ and ‘ii’.

Figures 3d and 4a: "Cycloneuralia" should be included in the phylogenetic trees.

OK, done.

Figure 3: The caption for the panel d is redundant. It should be changed into, for example, "Phylogenetic tree obtained from cladistic analyses using maximum likelihood (IQTREE)."

OK, done.

Supplementary Figures 6-9: In the captions, more detailed explanations of the results (for example, "50% majority rule consensus of XXX trees" and "strict consensus of all 4 most-parsimonious trees") should be provided.

OK, done.

Supplementary Figures 8 and 9: The caption explains that Cycloneuralia is resolved as a paraphyletic group, but it is not certain because Nematoida, Scalidophora, and Panarthropoda are resolved in a polytomy.

We changed the sentence into:

“Note that Cycloneuralia does not appear as a monophyletic clade”

Reviewer 3:

Line 25 'tiny' - I suggest giving an absolute measure of the size.

We add ‘maximal length 3 mm’.

Line 29 'both forms' - This is hard to follow by a non-expert. Can this be replaced with 'fossil species'?

OK, done.

Line 32 'dead-end' - Is this word necessary? I suggest to skip this word, as it is obvious that this lineage is extinct.

OK, done.

Lines 80, 94, and 172 'Remarks' - I, as a palaeontology non-expert, cannot get this manuscript structure with a repetition of this same section title.

Our systematic descriptions follow the standard rules in palaeontology.

Line 119 - I could not get what this 'Member 5' that was not introduced earlier means.

In Stratigraphy, ‘member’ is a lithostratigraphic subdivision (a Formation is usually subdivided into several Members).

Lines 104, 105, 417, ... - The name of the organization or database hosting these IDs (CUB.... and ELIXX....) should also be supplied.

OK, done.

Lines 341 and 361 - These two Figures (Figures 1 and 2) have the same caption (with an addition to the one for Figure 1). There should be a distinction based on what is presented in each figure.

We corrected the caption of Figure 2 and wrote the following: ‘Beretella spinosa gen. et sp. nov.’.

Line 362-367 - There is no guide about what the individual figure panels (e.g., Figure 2g, 2h, and 2i) show in detail. This guide should be supplied. This also applies to Figure 3a-c - are they anterolateral (a), dorsal (b), and posterolateral (c) views? It is better to write clearly in this way.

OK, done.

Figure 3d - The color contrast is not sufficient, and this figure does not look reader-friendly. Plus, the division into Cycloneuralia and Panarthropoda is indicated above the tree, but it is not clear what range of lineages these clades include. For example, is Pliciloricidae included in Cycloneuralia? Also, is Collinsium included in Panarthropoda? This figure looks quite unreliable, and it should be easy to fix.

OK, done.

Line 277 legend of Figure 3 - Including the parenthesis only with the program name (IQTREE) is not useful at all. Isn't it enough to describe it in Methods?

OK, done. We remove (IQTREE).

Line 380 legend of Figure 3 - I could not get where 'thicker bars' are.

Known fossil record indicated by thicker vertical bars. We added “vertical”.

Line 453 - Give full names of the methods, maximum parsimony, and maximum-likelihood.

OK, done.

Line 489 - State clearly what 'the recent paper' means.

Replace ‘recent’ by ‘present’.

Version 2

Abstract

Abstract

Extant ecdysozoans (moulting animals) are represented by a great variety of soft-bodied or articulated organisms that may or may not have appendages. However, controversies remain about the vermiform nature (i.e., elongated and tubular) of their ancestral body plan. We describe here Beretella spinosa gen et sp. nov. a tiny (maximal length 3 mm) ecdysozoan from the lowermost Cambrian, Yanjiahe Formation, South China, characterized by an unusual sack-like appearance, single opening, and spiny ornament. Beretella has no equivalent among animals, except Saccorhytus, also from the basal Cambrian. Phylogenetic analyses resolve both fossil species as a sister group (Saccorhytida) to all known Ecdysozoa, thus suggesting that ancestral ecdysozoans may have been non-vermiform animals. Saccorhytids are likely to represent an early off-shot along the stem-line Ecdysozoa. Although it became extinct during the Cambrian, this animal lineage provides precious insight into the early evolution of Ecdysozoa and the nature of the earliest representatives of the group.

Introduction

The Ediacaran‒Cambrian transition is marked by the appearance in the fossil record of a variety of new body plans that prefigure the majority of present-day animal lineages, including the ecdysozoans, a huge clade that encompasses all invertebrate animals growing through successive moulting stages, such as panarthropods (Arthropoda, Onychophora, Tardigrada), scalidophoran (incl. Priapulida) and nematoid worms (Erwin, 2020). Altogether ecdysozoans represent a very high percentage of animal biodiversity and disparity, inhabiting almost all possible ecological niches on Earth (Brusca et al., 2016). The nature of the last common ancestor of Ecdysozoa (LCAE) remains largely unresolved, even though worms are prevalent before the rise of panarthropods as trace and body fossils in basal Cambrian and late Ediacaran rocks (Buatois et al., 2014; Liu et al., 2014; Vannier et al., 2010). Some recent molecular phylogenies also predict that the most basal ecdysozoans were worm-like, elongated organisms (Howard et al., 2022; Laumer et al., 2019) that possibly diverged in the Ediacaran (Howard et al., 2022; Rota-Stabelli et al., 2013). Current reconstruction based on fossil and developmental evidence features the ancestral ecdysozoan as a millimeter-sized worm (Budd, 2001; Valentine and Collins, 2000) with a terminal (Ortega-Hernandez et al., 2019) or ventral mouth (Martín-Durán and Hejnol, 2015; Nielsen, 2019). Clearly, the discovery of Saccorhytus (Han et al., 2017; Liu et al., 2022; Shu and Han, 2020b) in the basal Cambrian of China (Kuanchuanpu Formation; ca. 535 Ma (Sawaki et al., 2008)) that is anything but a worm sowed doubt among scientists. Saccorhytus is a sac-like secondarily phosphatized microscopic animal spiked with conical sclerites and a single opening that was first seen as the earliest known deuterostome (Han et al., 2017) but is now considered as an ecdysozoan on more solid grounds (Liu et al., 2022; Shu and Han, 2020b), thus broadening the anatomical spectrum of the group and its disparity in the Cambrian and reopening the debate on the nature of LCAE.

We describe here Beretella spinosa gen. et sp. nov. from Member 5 of the Yanjiahe Formation (basal Cambrian Stage 2, ca. 529 Ma, Hubei Province, China) that shares morphological traits with Saccorhytus such as an ellipsoidal body, a pronounced bilaterality, a spiny ornament made of broad-based sclerites, and a single opening. Cladistic analyses are made to resolve the position of both Beretella and Saccorhytus that provide key information on the early evolution of the group.

Systematic palaeontology

Superphylum Ecdysozoa Aguinaldo et al. (Aguinaldo et al., 1997) Phylum Saccorhytida Han, Shu, Ou and Conway Morris, 2017 stat. nov.

Remarks

Saccorhytida first appeared in the literature as a new stem-group Deuterostomia that accommodated a single species, Saccorhytus coronarius (Han et al., 2017). Since Saccorhytus is no longer considered a primitive deuterostome and, instead, more likely belongs to ecdysozoans, Saccorhytida became an extinct Order of Ecdysozoa (Liu et al., 2022; Shu and Han, 2020b). Because both Saccorhytus and Beretella display major morphological differences with all other known ecdysozoan phyla (Nematoida, Scalidophora, and Panarthropoda), Saccorhytida is tentatively elevated here to the rank of phylum within Ecdysozoa.

Emended diagnosis

Microscopic, ellipsoidal body shape with pronounced bilateral symmetry expressed by paired spiny sclerites. Single, presumably oral opening on assumed ventral side (no anus).

Remarks

Only two forms, Saccorhytus and Beretella are currently placed within Saccorhytida, making it premature to formally define intermediate taxonomic categories such as an order and a family.

Beretella spinosa Han, Guo, Wang and Qiang, gen. et sp. nov.

LSID: urn:lsid:zoobank.org:act:C2DC9EC2-82EB-4B2B-9829-718EE8104593

Etymology

From ‘béret’, French, that designates a soft, visorless cap referring to the overall shape of this species, and ‘spinosa’, Latin, an adjective, alluding to its spiny ornament.

Holotype

CUBar138-12 (Fig. 1a‒c).

Beretella spinosa gen. et sp. nov. from Member 5 of the Yanjiahe Formation (Cambrian Stage 2), Yichang, Hubei Province, China.

a‒c, Holotype, CUBar138-12. a, Dorsal view showing the external ornament: (five sclerites at the midline in yellow (S1); flanked by two rows of sclerites in blue (S2); large broad-based conical sclerites in two dorsolateral pairs and one antero-posterior pairs in green (S3)); white arrows indicate lateral constriction. b, Right lateral view showing two additional rows of six sclerites (S4 and S5, in light blue and pink, respectively). c, Ventral view showing a large opening that may have accommodated the mouth (see the text) and an empty body cavity. d‒g, CUBar75-45. d, Dorsal view showing a broken S3. e, Micro-CT image, right lateral view displaying S4. f, Ventral view depicting a tiny projection in purple. g, An enlargement of the projection of f. h‒i, Paratype, CUBar171-5. h, Right dorsal view showing S1‒S4. i, Right-lateral view showing S4 and S5. j‒k, Paratype CUBar121-8. j, Dorsal view showing poorly preserved S1 and S2. k, Right-lateral view showing S3‒S5. A, assumed anterior end (see text); ef, exotic fragment; D, assumed dorsal side; L, left; P, posterior end; R, right; tp, tiny spine; V, ventral side. The same abbreviations are used throughout the manuscript including Supplementary materials.

Paratype

CUBar171-5 Fig. 1h, i) and CURBar121-8 (Fig. 1j, k).

Diagnosis

Body with a beret-like lateral profile. Convex side (presumably dorsal) with an elevated (presumably posterior) and lower (presumably anterior) end. The opposite side (presumably ventral) flattened. Bilateral symmetry well expressed in the overall body shape (sagittal plane) and sclerite distribution. Antero-posterior polarity. Convex side with a slightly elevated sagittal stripe topped with a single row of four aligned spines (S1) and five additional spines (S2) on each side. Six broad-based conical sclerites (S3) distributed in two symmetrical longitudinal rows plus two sagittal ones. Double rows of six marginal spines (S4 and S5). Flattened side often pushed in and partly missing, bearing a possible mouth opening. Possible oral spine.

Stratigraphy and locality

Watsonella crosbyi Assemblage Zone(Guo et al., 2021), Member 5 of the Yanjiahe Formation (Cambrian Terreneuvian, Stage 2) in the Yanjiahe section near Yichang City, Hubei Province, China (Supplementary Figs. 1, 2).

Beretella spinosa gen. et sp. nov.

a, CUBar99-19, dorsal view showing an ornament S1‒S4. b‒c, CUBar136-9, general dorsal view and details. d, CUBar136-11, dorsal view showing S1‒S5. e‒f, CUBar73-15 general view and details of the cuticular polygonal reticulation in black. g‒j, CUBar128-27. g, general view. h, i, details of outer and inner surface of the bi-layered structure of the cuticular wall as seen in broken conical sclerites. j, Micro-CT section showing possibly sclerite infilling. k-l, CUBar99-18, cuticular fragment, general view and details of large sclerite (central feature represents possible phosphatic infilling). is, infilling sclerite; pr, polygonal reticulation; sb, sclerite base.

Description and comparisons

The body of Beretella spinosa is secondarily phosphatized and has a consistent beret-like three-dimensional shape in the lateral view. Its length, width, and height range from 1000 to 2900 μm, 975 to 2450 μm, and 500 to 1000 μm, respectively (Fig. 1, Supplementary Tables 1‒3). The ratio of the maximal length to width is 1.6:1 (Supplementary Fig. 3). As seen in top view, B. spinosa shows a small lateral constriction at approximately mid-length (Fig. 1a, c).

Position of Beretella spinosa in the animal tree based on cladistic analysis.

a-c, artistic three-dimensional reconstructions of Beretella spinosa in the anterolateral (a), dorsal (b), and posterolateral views (c). d, Phylogenetic tree obtained from cladistic analyses using maximum likelihood. Saccorhytus and Beretella join in a clade (new phylum Saccorhytida) resolved as the sister-group of all other ecdysozoans; numbers at key nodes denote probability. Fossil and extant taxa are in italics and bold, respectively. Known fossil record indicated by thicker vertical bars (after (Shu and Han, 2020a)).

The body has a convex, assumedly dorsal side with one, presumably posterior end more elevated than the other (Fig. 1b, e, i, k). This elevation is gradual along the sagittal plane and then becomes more abrupt near the low elevated, presumably anterior end. The opposite, assumedly ventral side is less well preserved and seems to have been originally flattened.

The convex side bears a complex ornamented pattern made of five sets (S1‒S5) of spiny sclerites directed towards the more elevated end (Figs. 1a, b, d, e, h‒k, 2a, b, d). These sclerites were originally pointed (Figs. 1a, d, e, 2b, k, l, Supplementary Fig. 4a, b, g), but most of them were broken thus revealing an internal cavity and an ellipsoidal transverse section (Figs. 1a, b, h‒k, 2a‒e, g). The broken sclerites show an inner and outer phosphatic layer (thickness ca. 20 to 50 μm) often separated by a thin empty space (Fig. 2g‒l).

Possible evolutionary scenario to explain the origin and early evolution of ecdysozoans.

a, Summary tree (see Supplementary Figs. 6-9) showing saccorhytids as a sister-group of Cycloneuralia (Nematoida plus Scalidophora) + Panarthropoda; main morphological features of each group listed along each branch. b, Potential evolutionary pathway to evolve Saccorhytida and crown-group Ecdysozoa. Numbers in green, red and blue circles designate pre-ecdysozoan (Spiralia), Saccorhytida and Cycloneuralia, respectively. Light brown gradient (circle) to emphasize ecdysis and sclerite secretion seen as key evolutionary steps. 1, Hypothetical pre-ecdysozoan animal with a ciliated epidermis and glycocalyx. 2, Saccorhytid exemplified by Beretella with a cuticle bearing sclerites. 3, Crown-group ecdysozoan exemplified by a scalidophoran worm with an elongated shape, a differentiated head (introvert) and trunk, sclerites, a through gut, a terminal mouth and abilities to burrow into bottom sediment. Animals not to scale. Abbreviations: a, anus; a?, uncertain status of anus; ci, cilia; cu, cuticle; ec, epidermal cell; gl, glycocalyx (mucous layer); m, mouth; in, introvert; sc, sclerite; se, sediment; TGE, total-group Ecdysozoa. Silhouettes from phylopic.org.

The convex side bears six prominent conical sclerites (S3) all with a rounded to elliptical well-delimited broad base, distributed in two longitudinal symmetrical pairs with two additional sclerites at both ends of the sagittal plane (Figs. 1, 2d, Supplementary Figs. 3e‒h, 4e‒i). A low-relief stripe runs in a sagittal position and vanishes towards the elevated end. It is topped by a row of aligned spines (S1, Fig. 1a); the one closer to the more elevated end being more tubular and longer. This row is flanked on both sides by smaller aligned spines (S2, Figs. 1a, d, h, 2a‒c). Two relatively sinuous rows of six tiny spines are present parallel to the lateral margins (S4 and S5, Figs. 1b, e, h‒j, 2d, e). The convex side bears a polygonal micro-ornament (mesh size ca. 5 μm wide, Fig. 2f, Supplementary Tables 1‒3). However, its exact extension is uncertain due to coarse secondary phosphatization. Clusters of spherical phosphatized grains (diameter ca. 20 μm) occur near the sclerite base (Supplementary Fig. 4b).

In most specimens, the flattened side is occupied by a relatively large opening (1200 and 600μm in maximal length and width, respectively) with irregularly defined margins (Fig. 1c, f, see also supplementary information movies 1, 2). The flattened side is often largely missing and opens into a spacious internal cavity with no signs of internal organs (e.g. gut and pharynx) (Fig. 1c, f). One specimen shows a tiny spine on the margin of the flattened side (Fig. 1f, g), which differs from other spiny sclerites (S1-S5).

The length of studied specimens ranges from 1.0 to 2.9 mm (Supplementary Fig. 3e‒h). Whether growth was continuous or instead took place via successive moulting stages and cuticular renewal (ecdysis) could not be tested due to the small number of specimens (N=17) available for measurements. No major morphological variations (e.g., a sclerite pattern) can be seen between the smallest and largest specimens of B. spinosa (Supplementary Fig. 3e‒h).

Remarks

Body polarities in Beretella

The anterior-posterior (AP) and dorsal-ventral (DV) polarities of Beretella are uneasy to define because of the lack of modern equivalent among extant animals. In the vast majority of extinct and extant invertebrates for which antero-posterior polarity is defined on the basis of independent criteria (e.g. position of the mouth), sclerites point backwards (e.g. Cambrian scalidophoran worms (Han et al., 2007; Huang et al., 2004) and Wiwaxia (Zhang et al., 2015b)). This is most probably also the case with Beretella (Fig. 1a, d, j). The dorsoventral polarity of Beretella is supported by the fact that protective sclerites such as spines most commonly occur on the dorsal side of bilaterians (Fig. 1a, d, j).

Comparison with Saccorhytus and other ecdysozoans

Beretella spinosa has no exact equivalent in any Cambrian animals except Saccorhytus coronarius, an enigmatic, sac-like ecdysozoan (Han et al., 2017; Liu et al., 2022; Shu and Han, 2020b). Both forms share a tiny, poorly differentiated ellipsoidal body, and a set of prominent bilaterally arranged spiny sclerites. Indeed, the broad-based conical sclerites (S3) of Beretella are almost identical to those of Saccorhytus (Supplementary Fig. 4d) and have counterparts among scalidophoran worms (Supplementary Fig. 4c). However, they differ in number, ornamented structures, shape, and spatial arrangement. Beretella has a much more pronounced dorsoventral differentiation than Saccorhytus and its cuticle seems to have been harder and less flexible (see details in Supplementary Tables 1‒3), which altogether the hypothesis of Saccorhytus being the larval stage of Beretella unlikely. Both Beretella and Saccorhytus differ from other known ecdysozoans in the lack of an elongated body, introvert, annulations, and through gut (Supplementary Tables 1‒3).

Discussion

Ventral mouth

All bilaterian animals have a digestive system with at least one opening that corresponds to the mouth (Brusca et al., 2016). Although the presumed oral area of Beretella is poorly preserved (ventral side often pushed in and largely destroyed), its mouth is likely to be found ventrally (see description), since no other opening occurs on its dorsal side, except those created by broken sclerites. The well-defined dorsoventral polarity of Beretella would suggest that the animal was resting on its ventral (flattened) side, the spiny dorsal side playing a protective role.

Phylogenetic position of Beretella

Beretella’s phylogenetic affinities remain elusive due to the lack of information concerning its internal anatomy and ventral side. Its scleritome consists of isolated conical sclerites that were the cuticular outgrowths of a seemingly rigid integument that covered both sides of the animal. Such conical sclerites have close counterparts in Cambrian ecdysozoans such as scalidophoran worms (e. g. Eokinorhynchus (Zhang et al., 2015a)), lobopodians (e.g. Onychodictyon ferox (Hou et al., 1991)) and even more clearly Saccorhytus that recent cladistic analyses resolved as a branch of the total-group Ecdysozoa (Liu et al., 2022). These sclerites, unknown in other animal groups, suggest that both Saccorhytus and Beretella belongs to Ecdysozoa (moulting animals) although more direct fossil evidence such as exuviae or features suggesting cuticular moulting (Daley and Drage, 2016; Wang et al., 2019) are missing.

Cladistic analyses were performed to test the relation of Beretella and Saccorhytus to other ecdysozoan groups and, more generally, their phylogenetic relationships with other bilaterian groups (see details in Supplementary Figures 6-9). Both taxa join in a clade (Saccorhytida, Fig. 3a‒c) and are resolved as members of total-group Ecdysozoa. This clade is the sister group of Cycloneuralia plus Panarthropoda (crown-group Ecdysozoa, Figs. 3d, 4, Supplementary Figs. 6-9). These results are consistent with the body plan of Saccorhytida being markedly different from that of crown-group ecdysozoans that all have an elongated body and differentiated structures, such as, in Cycloneuralia, the introvert and pharyngeal complex (Fig. 4).

The ancestral ecdysozoan body plan

Molecular clock analyses often place the divergence of Ecdysozoa relatively deep into the Ediacaran (Howard et al., 2022; Rota-Stabelli et al., 2013), thus highlighting major discrepancy with the known fossil record of the group. Potential ecdysozoans occur in the late Precambrian as suggested by (i) sclerites resembling those of extant priapulids, found in Ediacaran Small Carbonaceous Fossils assemblages (Moczydłowska et al., 2015) and (ii) locomotion traces presumably made by scalidophoran worms (Buatois et al., 2014; Vannier et al., 2010). In the absence of fossil data for other groups such as nematoids, scalidophorans are potentially the oldest known representatives of Ecdysozoa. Recent Bayesian analyses based on a large molecular data set obtained from the 8 extant ecdysozoan phyla recover Scalidophora as the sister-group to Nematoida + Panarthropoda and suggest that ecdysozoans probably diverged in the Ediacaran, possibly some 23 million years before the oldest fossil occurrence (trace fossils) of the group (Howard et al., 2022). Although this study does not speculate on the nature of the last common ancestor of Ecdysozoa, it is consistent with the view that the earliest representatives of the group were probably worm-like, relatively elongated animals. (Howard et al., 2020) drew comparable conclusions based on Acosmia, an assumed stem-ecdysozoan worm from early Cambrian Chengjiang Lagerstätte. However, the re-evaluation of the morphological characteristics of this worm rather suggests a less basal position either within the total-group Cycloneuralia (Fig. 3d, Supplementary Figs. 6, 7) or among crown-group Ecdysozoa (Supplementary Figs. 8, 9). The ellipsoidal (non-vermiform) shape of saccorhytids and their position as the sister group of the crown-group Ecdysozoa clearly reopens the debate on the nature of the ancestral ecdysozoan (Fig. 4) and has led to explore alternative evolutionary hypotheses, in particular: (i) does the enigmatic saccorhytid body plan results from anatomical simplification? (ii) to what extent may these animals shed light on the nature of the earliest ecdysozoans?

Do saccorhytids result from simplification?

A relatively simple body plan and tiny size is often seen as resulting from anatomical simplification (e.g. reduction of digestive system) and miniaturization (micrometric size) in possible relation with the adaptation to specialized ecological niches or parasitism (Hanken and Wake, 1993). For example, some extant scalidophoran worms living in interstitial (meiobenthic) habitats such as loriciferans have a miniaturized body (Kirstensen, 1983) compared with their macroscopic counterparts (e.g. Priapulus (Schmidt-Rhaesa, 2013b)). However, they retain a through gut and a functional introvert and show no sign of drastic internal simplification (Schmidt-Rhaesa, 2013a). Anatomical reduction is a typical feature of parasitism(Hanken and Wake, 1993) that is well-represented among extant ecdysozoans such as nematodes (Schmidt-Rhaesa, 2014). Although relatively small (ca. 0.1-2.5 mm long), nematodes underwent no simplification of their digestive system. Saccorhytids have no specialized features (e.g. anchoring or piercing structures) that would point to any adaptation to ecto- or endo-parasitic lifestyles (Cong et al., 2017). Saccorhytus has been interpreted (Han et al. 2017) as a possible interstitial animal based on its micrometric size which corresponds to that of the extant meiofauna. If we accept the hypothesis that saccorhytids result from simplification, then we need to determine its origin. Simplification of saccorhytids from an elongated animal (e.g. cycloneuralian worm with a through gut and terminal mouth) is difficult to conceive because it would involve considerable anatomical transformations such as the loss of tubular organization, introvert and pharynx in addition to that of the digestive system (Fig. 4 and Supplementary Tables 4 and 5).

Early evolution of ecdysozoans: a new scenario

We propose here an alternative evolutionary hypothesis (Fig. 4) in which saccorhytids are replaced within the broader framework of the origin and early diversification of moulting animals. Saccorhytids are seen as an early off-shot from the stem-line Ecdysozoa (see cladistic analysis above) that possibly retained important features of the body plan of ancestral ecdysozoans. This scenario must be considered as a working hypothesis whose aim is to stimulate research in this key area of animal evolution.

The cuticular secretion and the loss of cilia (Valentine and Collins, 2000) would be the first of a series of evolutionary events (Fig. 4) that led to the rise of Ecdysozoa. Moulting (shedding of the old cuticle via apolysis and its renewal) reconciled body growth and cuticular protection (Schmidt-Rhaesa, 2007). Cuticle secretion and moulting may have been quasi-simultaneous innovations that took place over a relatively short time interval. The nature of the very first ecdysozoans is hypothetical and lacks fossil evidence. However, they are tentatively represented here as small epibenthic or interstitial slow-moving non-elongated animals from which saccorhytids may have evolved.

In our scenario, this ancestral ecdysozoan stock would have also given rise to elongated and tubular ecdysozoans through stepwise anatomical transformations such as the body elongation, the differentiation of key morpho-functional structures such as the pharynx and the introvert and the shift of the ventral mouth to a terminal position (Martín-Durán and Hejnol, 2015) (Fig. 4, and Supplementary Tables 4 and 5). This mouth shift from ventral to terminal arising in crown ecdysozoans is consistent with the chronology of divergence of animal lineages and the fact that the mouth of most spiralians is ventral (Martín-Durán and Hejnol, 2015; Nielsen, 2019; Ortega-Hernandez et al., 2019). Developmental studies show that embryos of extant cycloneuralians have a ventral mouth that moves to a terminal position towards the adult stage (Martín-Durán and Hejnol, 2015; Nielsen, 2019). These assumed major anatomical changes (e.g., functional introvert) must be placed in the ecological context of Cambrian animal radiation. Important changes in the functioning of marine ecosystems occurred in the early Cambrian such as interactive relationships between animal species, exemplified by predation (Vannier and Chen, 2005; Vermeij, 1977) and may have acted as drivers in the evolution of early ecdysozoans, in promoting burrowing into sediment and the colonization of endobenthic habitats for the first time (Vannier et al., 2010). Burrowing into the sediment could be seen as the evolutionary response of epibenthic animals such as ancestral ecdysozoans to escape visual predation (Daley et al., 2013; Vannier and Chen, 2005). We hypothesize that this migration to endobenthic shelters was made possible by the development of a resistant cuticular layer (Fig. 4) that strongly reduced physical damage caused by friction with the sediment and provided anchoring points (e.g. scalids and sclerites). Whereas saccorhytids became rapidly extinct during the Cambrian, worms massively colonized endobenthic habitats, resulting in bioturbation and ecological turnover.

Methods

Material

Fourteen specimens of Beretella spinosa were recovered from samples (siliceous-phosphatic, intraclastic limestone) collected from Member 5 of the Yanjiahe Formation, Yanjiahe section near Yichang City, Hubei Province, China (Guo et al., 2021) (Supplementary Tables 1‒3). These were obtained by digesting the rocks in 10% acetic acid. Faunal elements associated with Beretella spinosa (Supplementary Tables 1‒3) in residues are mainly tiny molluscs (CUBar21-4 and CUBar206-6). Comparisons were made with 10 specimens of Saccorhytus coronarius (ELIXX25-62, ELIXX34-298, ELIXX45-20, ELIXX48-64, ELIXX58-336, ELIXX61-27, ELIXX65-116, ELIXX65-296, ELIXX99-420) and one coeval scalidophoran specimen (ELIXX57-320) all from Bed 2 of the Kuanchuanpu Formation, Zhangjiagou section near Xixiang County, south Shaanxi Province, China. All specimens of Beretella are deposited in the paleontological collections of Chang’an University, Xi’an (CUBar), those of scalidophoran, and Saccorhytus at Northwest University, Xi’an (ELIXX), China.

Scanning electron microscopy (SEM)

All specimens were coated with gold and then imaged using a FEI Quanta 400 FEG SEM at Northwest University and a FEI Quanta 650 at Chang’an University.

X-ray computed microtomography and 3D reconstruction

Micro-CT-images (tiff format, with pixel size 1.1 μm) of Beretella (CUBar75-45, CUBar128-27, CUBar138-12) and Saccorhytus (ELIXX65-116, ELIXX99-420) were acquired using the Zeiss Xradia 520 at Northwest University (NWU), Xi’an, China, at an accelerating voltage of 50 kV and a beam current of 80 μA. Micro-CT data were processed using VGstudio Max 3.2 for 3D volume rendering.

Measurements

Measurements of the length, width, and height of Beretella and Saccorhytus were obtained from Micro-CT and SEM images by using tipDig2 v.2.16.

Phylogenetic analysis

We built our matrix with 55 taxa coded using 193 morphological characteristics (Supplementary Texts 1, 2). It is largely based on the data published by Howard et al. (Howard et al., 2020), Vinther and Parry (Vinther and Parry, 2019) and Ou et al. (Ou et al., 2017), although emended and supplemented by recent updates and new observations (Supplementary Text 1). Three characters (37. Through gut, 38. U-shaped gut, and 40. Ventral mouth) in matrix were coded as “? (uncertain)”, “?”, and “?”, respectively. Because although we can infer a ventral mouth and no anus of Beretella, these anatomic structures are invisible in fossils. We analyzed the data matrix using maximum parsimony (Tree analysis using New Technology, TNT), maximum likelihood (Important quartet tree, IQTREE) and Bayesian inference (MrBayes). Parsimony analysis was implemented in TNT under equal and implied (k=3) weight. Parameters are default (Goloboff et al., 2008; Goloboff and Catalano, 2016). The maximum-likelihood tree search was conducted in IQ-TREE (Nguyen et al., 2015), and support was assessed using the ultrafast phylogenetic bootstrap replication method (Hoang et al., 2018; Minh et al., 2013) to run 50,000 replicates. Bayesian inference was conducted in with MrBayes v3.2.6a with default priors and Markov chain Monte Carlo settings(Ronquist et al., 2012). Two independent runs of 7,000,000 Markov chain Monte Carlo generations were performed, each containing four Markov chains under the Mkv + Γ model for the discrete morphological character data (Lewis, 2001). In each run (N=2), trees were collected at a sampling frequency of every 5,000 generations and with the first 25% samples discarded as burn-in. The convergence of chains was checked by effective sample size (ESS) values over 1,000 in Tracer v.1.7 (Rambaut et al., 2018), 1.0 for the potential scale reduction factor (PSRF)(Gelman and Rubin, 1992), and by an average standard deviation of split frequencies below 0.007.

Ancestral character state reconstructions

Ancestral character state reconstructions for four morphological characters were performed on the ecdysozoan total group node, the ecdysozoan crown group node and saccorhytid node. Cycloneuralia was treated as (i) a monophyletic (Supplementary Table 4) and (ii) paraphyletic group (Supplementary Table 5). Characters selected for ancestral state reconstruction represent traits inferred as ecdysozoan plesiomorphies (ancestral characters) from studies of crown group taxa. These characters included the presence or absence of: (1) through gut; (2) ventral mouth; (3) introvert (see Supplementary Tables 4 and 5).

This was carried out individually for the selected character in MrBayes. This was employed to calculate the posterior probability of the presence (1) and absence (0) of the selected characters at the selected nodes. Analyses used the MK + gamma model, and always converged after 2 million generations. Average deviation of split frequencies (< 0.01), ESS scores (> 200), and PSRF values (= approx. 1.00) assessed convergence of the MCMC chains (Howard et al., 2020).